---
title: "Palatal consonants in PHOIBLE"
author: "Steven Moran\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
      toc: true
bibliography: 'references.bib'
---

# Overview

This report does tasks like extracting [palatal consonants](https://en.wikipedia.org/wiki/Palatal_consonant) from [PHOIBLE](https://phoible.org/) [@MoranMcCloy2019].


## Data preparation

This report uses these [R](https://cran.r-project.org/) libraries [@tidyverse;@knitr]:

```{r, message=F, warning=F}
library(tidyverse)
library(knitr)
```

Get the PHOIBLE data and merge in the metadata from [Glottolog](https://glottolog.org/) [@Hammarstrom_etal2020].

PHOIBLE data is released periodically be version because the editors add new data, make reported corrections, etc. Here we use the PHOIBLE dev version from [May 2020](https://github.com/phoible/dev/tree/646f5e4f64bfefb7868bf4a3b65bcd1da243976a).

```{r}
phoible <- read_csv(url("https://github.com/phoible/dev/blob/646f5e4f64bfefb7868bf4a3b65bcd1da243976a/data/phoible.csv?raw=true"), 
                    col_types = c(InventoryID = "i", Marginal = "l", .default = "c"))
```

Merge in [Glottolog 4.1](https://glottolog.org/meta/downloads) data.

```{r}
languoids <- read.csv("glottolog_languoid.csv/languoid.csv", stringsAsFactors = FALSE)
geo <- read.csv("languages_and_dialects_geo.csv", stringsAsFactors = FALSE)
phoible <- left_join(phoible, languoids, by = c("Glottocode" = "id"))
phoible <- left_join(phoible, geo)
rm(geo, languoids)
```

Create a PHOIBLE index.

```{r}
index <- phoible %>%
  select(InventoryID, 
         Glottocode, 
         ISO6393, 
         name, 
         LanguageName, 
         SpecificDialect, 
         Source, 
         family_id, 
         level, 
         status, 
         latitude, 
         longitude, 
         country_ids, 
         macroarea) %>%
    distinct()

index <- index %>% rename(GlottologName = name, PhoibleName = LanguageName)
```

Get palatal consonant counts and their number of marginals.

```{r}
palatals <- phoible %>% filter(grepl("j|ɲ|c|ɟ|ç|ʝ|ʎ", Phoneme))

palatal_counts <- palatals %>%
  group_by(InventoryID) %>%
  summarize(palatals = n())

palatals_marginals <- palatals %>%
  filter(Marginal) %>%
  group_by(InventoryID) %>%
  summarize(Marginal_palatals = n())
```
Let's write the palatals count data to disk. The output data format is a [CSV](https://en.wikipedia.org/wiki/Comma-separated_values) file with the following columns:

```
InventoryID | Glottocode | ... | Palatals | Marginal_palatals | ...
1 | aaaa1111 | ... |
```

```{r}
write_csv(palatals, 'palatals.csv')
```


## Exploratory

Merge together the palatal and marginal palatal counts.

```{r}
df <- left_join(palatal_counts, palatals_marginals)
```

Have a look at the data.

```{r}
df %>%
  head() %>%
  kable()
```

Do any languages contain palatals that are always marginal? Yes.

```{r}
temp <- df[which(df$palatals == df$Marginal_palatals), ]
temp %>% kable()
```

Which languages and which palatals are in that set?

```{r}
phoible %>% select(InventoryID, Glottocode, LanguageName, Phoneme) %>% filter(InventoryID %in% temp$InventoryID) %>% filter(grepl("j|ɲ|c|ɟ|ç|ʝ|ʎ", Phoneme))
```

How many Glottocodes are there in phoible?

```{r}
nrow(phoible %>% select(Glottocode) %>% distinct())
```

How many phoible inventories have palatal consonants?

```{r}
nrow(palatals %>% select(InventoryID) %>% distinct())
```

How many are marginal?

```{r}
nrow(palatals %>% filter(Marginal) %>% group_by(InventoryID, Marginal)) # 70 rows
kable(palatals %>% filter(Marginal) %>% group_by(InventoryID, Marginal))
```

How are palatals distributed in phoible (across duplicate languages)?

```{r}
distribution_palatals <- palatals %>%
  group_by(Phoneme) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
kable(distribution_palatals)

# Order the frequency counts and plot the distribution of uvular consonants in the sample
distribution_palatals$Phoneme <- factor(distribution_palatals$Phoneme, levels = distribution_palatals$Phoneme[order(-distribution_palatals$count)])

# qplot(distribution_palatals$Phoneme, distribution_palatals$count)

p <- ggplot(aes(y = count, x = Phoneme), data = distribution_palatals) +
  geom_bar(stat = "identity", width = 0.3, color = "black") +
  xlab("Segments") +
  ylab("# of languages") +
  theme_minimal() +
  ggtitle("")
p
```


Distribution of palatal consonants per inventory (can't use Glottocode because there are multiple doculects).

```{r}
palatals_counts <- palatals %>%
  select(InventoryID, Glottocode, Phoneme, macroarea) %>%
  group_by(InventoryID, Glottocode, macroarea) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

qplot(y = palatals_counts$count)
```

How are they distributed via macroarea?

```{r}
kable(table(palatals_counts$macroarea))
```


## Some searches

For example, languages that have /ʎ/ and /c/ simultaneously.

```{r}
temp <- phoible %>% filter(Phoneme %in% c('ʎ', 'c'))
tallied <- temp %>% group_by(InventoryID) %>% tally()
tallied <- tallied %>% filter(n>1)
temp %>% filter(InventoryID %in% tallied$InventoryID) %>% arrange(InventoryID) %>% kable()
```


# References
