---
title: "Identify languages of Ethiopia in PHOIBLE"
author: "Steven Moran"
output:
  github_document:
  pandoc_args: --webtex
---

In regard to [issue 299](https://github.com/phoible/dev/issues/299), let's check the status of certain segments in Ethiopian languages.


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(testthat)
```

Load phoible.

```{r}
col_types <- cols(InventoryID='i', Marginal='l', .default='c')
phoible <- read_csv(url('https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true'), col_types = col_types)
```

Load Glottolog.

```{r, warning=FALSE, message=FALSE}
glottolog <- read_csv('glottolog_languoid.csv/languoid.csv')
```

Subset phoible.

```{r}
cut <- phoible %>% select(InventoryID, LanguageName, Source, Glottocode) %>% distinct()
```

Merge in Glottlog.

```{r}
cut <- left_join(cut, glottolog, by=c("Glottocode"="id"))
```

Subset languages reportedly spoken in Ethiopia.

```{r}
cut <- cut %>% filter(grepl("ET", country_ids)) %>% arrange(InventoryID)
```

There are 72 inventories.

```{r}
cut %>% kable()
```

Sascha Völlmin notes that in [Zay](https://phoible.org/inventories/view/1510), [q] should be [k'] (the traditional notation for ejective k' in Ethiopistic studies).

Let's see which of these languages have [q] in the inventories. First get the Ethiopian inventories.

```{r}
e_inventories <- phoible %>% filter(InventoryID %in% cut$InventoryID)

# Double check that we have the same number of inventories
expect_equal(nrow(e_inventories %>% select(InventoryID) %>% distinct()), nrow(cut))
```

Then select the ones with [q].

```{r}
e_inventories %>% filter(grepl("q", Phoneme)) %>% select(InventoryID, Glottocode, LanguageName, SpecificDialect, Source, Phoneme, Marginal) %>% kable()
```

I suspect that the sources `spa` and `upsid` are correct, e.g. Awngi (here listed with an inapproriate ethnonym) and Hamar indeed have a uvular [q]:

* https://en.wikipedia.org/wiki/Awngi_language
* https://en.wikipedia.org/wiki/Hamer_language

And that the source `gm` inventories are potentially incorrect, including Zway. Fulfulde is spoken mainly in West Africa:

* https://en.wikipedia.org/wiki/Fula_language

But we should check the status of [q] there.

Agaw's glottocode:

* https://glottolog.org/resource/languoid/id/qima1242

notes that the dialect is Qimant: 

https://en.wikipedia.org/wiki/Qimant_language

We should also check these sources since both Fula and Qimant seem to lack uvular [q].

The rest of the `gm` languages in Ethiopia need to be checked for systematic transcriptio errors for [q] and any other transcription practices in Ethiopistic studies.

