---
title: "Papuan phoneme inventories in PHOIBLE"
author: ""
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document
---

Load the R libraries.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(maps)
```

Get the current [PHOIBLE inventories](https://phoible.org) in [CLDF format](https://github.com/cldf-datasets/phoible). See also:

* https://cldf.clld.org
* https://clld.org/datasets.html


```{r, warning=FALSE, message=FALSE}
values <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/phoible/master/cldf/values.csv'))
languages <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/phoible/master/cldf/languages.csv'))
inventories <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/phoible/master/cldf/inventories.csv'))
parameters <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/phoible/master/cldf/parameters.csv'))
```

Merge CLDF tables into a single data frame at the level of inventories.

```{r, warning=FALSE, message=FALSE}
phoible <- values %>% select(Language_ID, Inventory_ID) %>% distinct()
phoible <- left_join(phoible, inventories, by=c("Inventory_ID" = "ID"))
phoible <- phoible %>% rename(Source_name = Name)
phoible <- left_join(phoible, languages, by=c("Language_ID" = "ID"))
```

This is what the data format looks like.

```{r}
phoible %>% head()
```

Merge CLDF tables into a single data frame at the level of segments.

```{r}
segments <- left_join(values, inventories, by=c("Inventory_ID" = "ID"))
segments <- segments %>% rename(Source_name = Name)
segments <- left_join(segments, languages, by=c("Language_ID" = "ID"))
parameters <- parameters %>% rename(SegmentName = Name)
segments <- left_join(segments, parameters, by=c("Parameter_ID" = "ID"))
```

This is how the data look at the level of segments.

```{r}
segments %>% filter(Name == "Hawaiian") %>% head()
```


Filter out languages -- note this is not 100% correct according to Harald Hammarstrom (p.c.), i.e., Glottolog Papunesian languages minus Austronesian does not equal Papuan languages exactly, so this is an approximation for exploratory analysis.

```{r}
png <- segments %>% filter(Macroarea == "Papunesia")
png <- png %>% filter(Family_Name != "Austronesian")
```

Let's check those data points on a world map.

```{r}
ggplot(data=png, aes(x=Longitude,y=Latitude)) + 
  borders("world", colour="gray50", fill="gray50") + 
  geom_point()
```

Tag the languages as Papuan in PHOIBLE and compare them with the worldwide sample for segment inventory sizes. Note that this is non-inclusive, i.e., in the whole sample a language is either Papuan or not (according to the definition above). This means we are comparing "only" Papuan languages versus non-Papuan languages in the sample (instead of Papuan versus the whole sample inclusive of Papuan languages).

```{r}
phoible <- phoible %>% mutate(Papuan = ifelse(Language_ID %in% png$Language_ID, T, F))
```

Make some plots.

Total phonemes.

```{r}
phoible %>%
ggplot(aes(x = Papuan, y = count_phonemes)) +
  geom_boxplot() + 
  theme_bw()

phoible %>% filter(count_phonemes <= 90) %>%
ggplot(aes(x = Papuan, y = count_phonemes)) +
  geom_boxplot() + 
  theme_bw()

phoible %>% filter(count_phonemes <= 50) %>%
ggplot(aes(x = Papuan, y = count_phonemes)) +
  geom_boxplot() + 
  theme_bw()
```

Consonants.

```{r}
phoible %>%
ggplot(aes(x = Papuan, y = count_consonants)) +
  geom_boxplot() + 
  theme_bw()

phoible %>% filter(count_consonants <= 90) %>%
ggplot(aes(x = Papuan, y = count_consonants)) +
  geom_boxplot() + 
  theme_bw()

phoible %>% filter(count_consonants <= 50) %>%
ggplot(aes(x = Papuan, y = count_consonants)) +
  geom_boxplot() + 
  theme_bw()
```

Vowels.

```{r}
phoible %>%
ggplot(aes(x = Papuan, y = count_vowels)) +
  geom_boxplot() + 
  theme_bw()

phoible %>% filter(count_vowels <= 30) %>%
ggplot(aes(x = Papuan, y = count_vowels)) +
  geom_boxplot() + 
  theme_bw()
```

Tones.

```{r}
phoible %>% filter(count_tones >= 1) %>%
ggplot(aes(x = Papuan, y = count_tones)) +
  geom_boxplot() + 
  theme_bw()
```