---
title: "Which languages in PHOIBLE have multiple inventories?"
author: "Steven Moran"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(testthat)
library(dplyr)
library(knitr)
```

PHOIBLE has both ISO 639-3 language name identifiers (aka language codes) and Glottocodes for most phonological inventories in its dataset:

https://github.com/phoible/dev/blob/master/mappings/InventoryID-LanguageCodes.csv

ISO codes identify language-level entries. Glottocodes identify all languoids, i.e. all families, languages, and dialects. In PHOIBLE, we assign the "lowest" level Glottocode, e.g. the dialect level when it's known for a given phonological inventory.

This means that there are two levels of multiple inventories for data points in PHOIBLE: at the ISO versus the Glottocode levels.

```{r}
# Get the latest phoible dev data
load(url('https://github.com/phoible/dev/blob/master/data/phoible.RData?raw=true'))
expect_equal(nrow(phoible), 105467) # latest phoible number of data points
```

```{r}
# Collapse inventories and get each inventory's phoneme count
counts <- phoible %>% group_by(InventoryID, Glottocode, ISO6393) %>% summarize(phonemes = n()) %>% arrange(ISO6393, Glottocode, phonemes)
expect_equal(nrow(counts), 3020) # phoible 2.0 has 3020 data points
```

```{r}
# Get the distribution of multiple inventories at the ISO level
dup.iso <- counts %>% group_by(ISO6393) %>% summarize(inventories = n()) %>% arrange(desc(inventories))
expect_equal(nrow(dup.iso), 2099) # phoible 2.0 number of unique iso codes (should be less than dup glottocodes)

# What's the distribution?
kable(table(dup.iso$inventories))
```

```{r}
# Get the distribution of multiple inventories at the Glottocode level
dup.glottocodes <- counts %>% group_by(Glottocode) %>% summarize(inventories = n()) %>% arrange(desc(inventories))
expect_equal(nrow(dup.glottocodes), 2184) # phoible 2.0 number of unique glottocodes

# What's the distribution?
kable(table(dup.glottocodes$inventories))
```

```{r}
# Show some ISO singletons
single.iso <- dup.iso %>% filter(inventories == 1)
temp <- left_join(counts, single.iso)
kable(temp %>% arrange(InventoryID) %>% filter(!is.na(inventories)) %>% head())
```



```{r}
# Do the ranges, means, and stuff for ISO
temp1 <- counts %>% group_by(ISO6393) %>% summarize(num_inventories = n(), min_phonemes = min(phonemes), max_phonemes = max(phonemes), mean_phonemes = mean(phonemes)) %>% arrange(desc(num_inventories))
temp2 <- counts %>% group_by(ISO6393) %>% summarize(range = paste(phonemes, collapse=","))
temp <- left_join(temp1, temp2)
```

Note range may not be ordered correctly.

```{r}
kable(temp %>% head(n=20))
```

```{r}
# Do the ranges, means, and stuff for Glottocodes
temp1 <- counts %>% group_by(Glottocode) %>% summarize(num_inventories = n(), min_phonemes = min(phonemes), max_phonemes = max(phonemes), mean_phonemes = mean(phonemes)) %>% arrange(desc(num_inventories))
temp2 <- counts %>% group_by(Glottocode) %>% summarize(range = paste(phonemes, collapse=","))
temp <- left_join(temp1, temp2)
```

Note range may not be ordered correctly.

```{r}
kable(temp %>% head(n=20))
```

```{r}
# Clean up
rm(list = ls())
```