---
title: "Which languages in PHOIBLE have multiple inventories?"
author: "Steven Moran"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
library(testthat)
library(dplyr)
library(knitr)
```

# Overview

PHOIBLE has both ISO 639-3 language name identifiers (aka language codes) and Glottocodes for most phonological inventories in its dataset:

https://github.com/phoible/dev/blob/master/mappings/InventoryID-LanguageCodes.csv

ISO codes identify language-level entries. Glottocodes identify all languoids, i.e. all families, languages, and dialects. In PHOIBLE, we assign the "lowest" level Glottocode, e.g. the dialect level when it's known for a given phonological inventory.

This means that there are two levels of multiple inventories for data points in PHOIBLE: at the ISO versus the Glottocode levels.

Get the latest phoible dev data.

```{r}
phoible <- read.csv('https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true')
# expect_equal(nrow(phoible), 105467) # latest phoible number of data points
```

Collapse inventories and get each inventory's phoneme count (note the warning that some have NAs).

```{r}
counts <- phoible %>% group_by(InventoryID, Glottocode, ISO6393) %>% summarize(phonemes = n()) %>% arrange(ISO6393, Glottocode, phonemes)
expect_equal(nrow(counts), 3020) # phoible 2.0 has 3020 data points
```

Get the distribution of multiple inventories at the ISO level.

```{r}
dup.iso <- counts %>% group_by(ISO6393) %>% summarize(inventories = n()) %>% arrange(desc(inventories))
# expect_equal(nrow(dup.iso), 2092) # phoible 2.0 number of unique iso codes (should be less than dup glottocodes)
```

We expect more duplicates entries at the ISO 639-3 level because it encodes languages, but phoible often has multiple data points at the dialect level.

```{r}
table(dup.iso$inventories) %>% kable()
```

Note that some ISO 639-3 codes are (currently) NA, hence the 37 "duplicates" in the tables above and below.

```{r}
head(dup.iso) %>% kable()
```

Get the distribution of multiple inventories at the Glottocode level.

```{r}
dup.glottocodes <- counts %>% group_by(Glottocode) %>% summarize(inventories = n()) %>% arrange(desc(inventories))
# expect_equal(nrow(dup.glottocodes), 2184) # phoible 2.0 number of unique glottocodes
```

What's the distribution?

```{r}
table(dup.glottocodes$inventories) %>% kable()
```

There are less duplicates per Glottocode code (subsumes dialects).

```{r}
head(dup.glottocodes) %>% kable()
```

Here's an example.

```{r}
counts %>% filter(Glottocode=="osse1243") %>% arrange(InventoryID)
```

```{r}
counts %>% filter(ISO6393=="oss") %>% arrange(InventoryID)
```

Show some ISO singletons.

```{r}
single.iso <- dup.iso %>% filter(inventories == 1)
temp <- left_join(counts, single.iso)
temp %>% arrange(InventoryID) %>% filter(!is.na(inventories)) %>% head() %>% kable()
```

Get the ranges, means, and stuff for ISO.

```{r}
temp1 <- counts %>% group_by(ISO6393) %>% summarize(num_inventories = n(), min_phonemes = min(phonemes), max_phonemes = max(phonemes), mean_phonemes = mean(phonemes)) %>% arrange(desc(num_inventories))
temp2 <- counts %>% group_by(ISO6393) %>% summarize(range = paste(phonemes, collapse=","))
temp <- left_join(temp1, temp2)
```

Note range may not be ordered correctly. Again, NA is meaningless here.

```{r}
temp %>% head(n=20) %>% kable()
```

Get the ranges, means, and stuff for Glottocodes.

```{r}
temp1 <- counts %>% group_by(Glottocode) %>% summarize(num_inventories = n(), min_phonemes = min(phonemes), max_phonemes = max(phonemes), mean_phonemes = mean(phonemes)) %>% arrange(desc(num_inventories))
temp2 <- counts %>% group_by(Glottocode) %>% summarize(range = paste(phonemes, collapse=","))
temp <- left_join(temp1, temp2)
```

Note range may not be ordered correctly. Some languages have consistently the same (or similar) numbers of phonemes. Others vary greatly.

```{r}
temp %>% head(n=20) %>% kable()
```

# Compare differences individual inventories

We can use the `pull` command in `dplyr` to compare phoneme inventories next to each other. For example, incorrect ejectives in one of the two barb1263 inventories.

## Example of doculect differences

### barb1263

Barbareño (https://glottolog.org/resource/languoid/id/barb1263) is an example of two inventories from different doculects (see references below) that have slightly different phoneme inventories.

https://phoible.org/inventories/view/1239

versus:

https://phoible.org/inventories/view/862

Inventory (PH) 1239 has 44 phonemes:

```{r}
phoible %>% filter(InventoryID == 1239) %>% pull(Phoneme)
```

Inventory (PH) 862 has 40 phonemes:

```{r}
phoible %>% filter(InventoryID == 862) %>% pull(Phoneme)
```

The difference between the two inventories is reflected clearly in their sources, i.e. Beeler 1971 posits four additional segments (/ s̰, ̰ʃ,̰ x, ʃʰ /)whereas Wash 2001 does not.

References:

``` 
@phdthesis{boi_wash2001,
	Author = {Wash, Suzanne},
	Date-Added = {2013-09-01 18:16:42 +0000},
	Date-Modified = {2013-09-01 18:16:42 +0000},
	Filenames = {boi_wash2001.pdf},
	School = {The University of California at Santa Barbara},
	Title = {{Adverbial Clauses in Barbare{\~n}o Chumash Narrative Discourse}},
	Year = {2001}}

@article{boi_chumash1970,
	Author = {Beeler, M. S.},
	Date-Added = {2013-09-01 18:16:42 +0000},
	Date-Modified = {2013-09-01 18:16:42 +0000},
	Filenames = {boi_chumash1970.pdf},
	Journal = {International Journal of American Linguistics},
	Number = {1},
	Pages = {14--17},
	Publisher = {The University of Chicago Press},
	Title = {{Sibilant Harmony in Chumash}},
	Volume = {36},
	Year = {1970}}
```


## trum1247

Two inventories for Trumai (https://glottolog.org/resource/languoid/id/trum1247), an isolate spoken in Brazil, are given as follows. The first comes from UPSID (Maddieson 194l Maddieson & Precoda 1990) and the second from work by Guiardello (1999). This is another case where more recent fieldwork reanalyzes the language. (Note also that inventories in UPSID were typologized by Maddieson, i.e. sometimes reanalyzed so that the segments were unified across descriptions to make typological comparison possible. The work by Maddieson built on the precursor database, the Stanford Phonology Archive, and there are systematic differences between inventories in the two sources from the same doculects. We revisit this issue in a separate section. )

https://phoible.org/inventories/view/588

```{r}
phoible %>% filter(InventoryID == 588) %>% pull(Phoneme)
```

Note it does not contain ejectives. The second is from Guiardello (1999) and it does list ejectives. 

https://phoible.org/inventories/view/1939

```{r}
phoible %>% filter(InventoryID == 1939) %>% pull(Phoneme)
```

Guiardello (1999:1) writes:

```
Trumai has 23 consonants and 6 vowels. In Guiardello (1992), the number of Trumai phonemic consonants was smaller, given that the analysis for some phones was different from the current one: the lateral fricative /ɬ/ was analyzed as an allophone of /l/, but now it is analyzed as a independent phoneme; the affricate /ts/ and the ejectives /t̪ʼ/, /tʼ/,/kʼ/, and /tsʼ/ were previously classified as consonant clusters (i.e. /t+s/ and /t+s+ʔ/), but now we believe they are more adequately analyzed as single phonemes. Therefore, the consonant chart presented here has some differences in relation to the one presented in previous work. 

```

Wikipedia (https://en.wikipedia.org/wiki/Trumai_language) notes:

```
This inventory is atypical of Amazonian languages (Trumai is a recent immigrant to the Xingu basin) in its ejective consonants, the lateral fricative /ɬ/, and the alveolar–dental distinction. Guirardello, who specializes on Trumai, has presented varied inventories of these phonemes: Guirardello (1999a)[4] lists /t̪ t̪' ts ts' s/ as dental, and /t t' d n l ɬ ɾ/ as alveolar; whereas Guirardello (1999b) lists only /t/ and /t'/ as alveolar.[18] Younger speakers do not make the ejective distinction.
```

```
@book{Trumai1975,
	Address = {Paris},
	Author = {Monod-Becquelin, Aurore},
	Date-Added = {2013-09-01 12:19:49 +0000},
	Date-Modified = {2013-09-01 12:19:49 +0000},
	Iso_Code = {tpy},
	Keywords = {Trumai},
	Olac_Field = {phonology},
	Publisher = {Centre National de la Recherche Scientifique},
	Refdb_Id = {http://wals.info/refdb/record/575},
	Title = {{La pratique linguistique des indiens trumai (Haut-Xingu, Mato Grosso, Br{\'{e}}sil)}},
	Wals_Code = {tru},
	Year = {1975}}

@phdthesis{1939_Guiardello1999,
	Author = {Guiardello, Raquel},
	Bibtexkey = {Guiardello1999},
	Date-Added = {2014-07-09 13:41:25 +0000},
	Date-Modified = {2014-07-09 13:41:26 +0000},
	Inventoryid = {1939},
	Languagecode = {tpy},
	Languagename = {Trumai},
	Languagevariantcode = {tpy},
	Pages = {1--6},
	School = {Rice -- University},
	Title = {A Reference Grammar of Trumai},
	Year = {1999}}
```

### Khanty

https://phoible.org/languages/khan1273



### Greek

https://phoible.org/languages/mode1248




















# TODO

### Moseten


```{r}
phoible %>% filter(InventoryID == 944) %>% pull(Phoneme)
```

```{r}
phoible %>% filter(InventoryID == 1986) %>% pull(Phoneme)
```


## Uvulars


## Same doculect, same inventory?

We need to check that inventories from the same source documents contain the same realizations, e.g.

* https://phoible.org/inventories/view/2266
* https://phoible.org/inventories/view/2219

```{r}
phoible %>% filter(InventoryID == 2266) %>% pull(Phoneme)
```

```{r}
phoible %>% filter(InventoryID == 2219) %>% pull(Phoneme)
```


## Investigating Dan

* https://phoible.org/inventories/view/298
* https://phoible.org/inventories/view/1393


```{r}
phoible %>% 
  filter(InventoryID == 1393 | InventoryID == 298) %>%
  select(InventoryID, Phoneme) %>%
  group_split(InventoryID) %>%
  lapply(function(tib) pull(tib, Phoneme))
```



```{r}
phoible %>% 
  filter(InventoryID == 962 | InventoryID == 1411) %>%
  select(InventoryID, Phoneme) %>%
  arrange(Phoneme) %>%
  group_split(InventoryID) %>%
  lapply(function(tib) pull(tib, Phoneme))
```

## SPA vs UPSID

Maddieson typologized many of the phonological inventories in SPA when creating UPSID. Hence, many of the inventories point to the same bibliographic sources. Moran (2012) notes some systematic difference between the UPSID and SPA, in particular, the way in which the vowels are encoded with diacritics in UPSID. UPSID also jettisoned descriptions of tone.

Investigate Cham in SPA vs UPSID (todo: make a comparison of all SPA/UPSID languages and quantify the differences).

```{r}
# https://github.com/phoible/dev/issues/276
# 
```




```{r}
# Clean up
rm(list = ls())
```