---
title: "Advanced front vowel?"
author: "Steven Moran \\<steven.moran@unine.ch\\>"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---
  
```{r, message=F, warning=F}
library(tidyverse)
library(knitr)
```

# Overview

A question regarding [a̟]: as an advanced front vowel is raised here:

https://linguistics.stackexchange.com/questions/37345/a%cc%9f-advanced-front-vowel


So in which inventory/ies does it occur and why?

Load the PHOIBLE development data from the GitHub repository.

```{r}
phoible <- read_csv(url('https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true'), col_types=c(InventoryID='i', Marginal='l', .default='c'))
```

Get just the rows that contain an advanced [a].

```{r}
advanced_a <- phoible %>% filter(grepl("a̟", Phoneme))
```

Trim it down a bit to check if it's something that's source-specific.

```{r}
advanced_a %>% select(InventoryID, Glottocode, ISO6393, LanguageName, Source, Phoneme) %>% kable()
```

It's mainly in SPA and UPSID (with two data points in EA).

## EA

[EA](http://eurasianphonology.info/) lists all its segments [here](http://eurasianphonology.info/segments). In the "Plain series" for vowels, advanced a is listed under "near-front central". It is found in [Saami](http://eurasianphonology.info/search_exact?dialects=True&query=a%CC%9F)

In the [pre-release version of EA](https://eurphon.info/), [Catalan (Mallorcan)](https://eurphon.info/languages/html?lang_id=443) also had the advanced [a]. The inventory seems to have been removed in the stable version of EA. The inventory source:

* Lloret, Maria-Rosa. 2011. La fonologia del català. Col. «El que sabem de...» Articles de suport a la docència. Barcelona: Santillana.


## UPSID

In UPSID, the conversion from ASCII representations to Unicode IPA shows that:

* [low front unrounded vowel](https://github.com/phoible/dev/blob/master/raw-data/UPSID/UPSID_IPA_correspondences.tsv#L819) is encoded as <a+>

and

* [low central unrounded vowel](https://github.com/phoible/dev/blob/master/raw-data/UPSID/UPSID_IPA_correspondences.tsv#L809) as <a>

The low front unrounded vowel is reported in 26 languages (our of 451):

* http://menzerath.phonetik.uni-frankfurt.de/S/S0302.html


## SPA

The [IPA correspondences in SPA](https://github.com/phoible/dev/blob/master/raw-data/SPA/SPA_IPA_correspondences.tsv) are more detailed than UPSID because SPA contains a rich set of allophonic (phonetic) descriptions.

Here we also have an "a" vs "a-front" distinction:

* https://github.com/phoible/dev/blob/master/raw-data/SPA/SPA_IPA_correspondences.tsv#L1536-L1537


## Contrastive?

Let's look and see if there is an "a" vs "a-front" phonemic contrast in any of these languages.

```{r}
df <- phoible %>% filter(InventoryID %in% advanced_a$InventoryID) %>% select(InventoryID, Glottocode, ISO6393, LanguageName, SpecificDialect, Source, Phoneme, Allophones, SegmentClass)
df %>% filter(grepl("a", Phoneme)) %>% kable()
```

And indeed there is -- although not in every case of "a" vs "a-advanced".







