---
title: "Check for errors in PHOIBLE segment conventions"
author: "Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
  pandoc_args: --webtex
---
  
Let's load the libraries and PHOIBLE.
  
```{r, message=F, warning=F}
library(dplyr)
library(readr)
library(testthat)
library(knitr)
```

```{r}
col_types <- cols(InventoryID = "i", Marginal = "l", .default = "c")
phoible <- read_csv(url("https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true"), col_types = col_types)

# Note the phoible dev version
expect_equal(nrow(phoible), 105460) # https://github.com/phoible/dev/commit/9d21f8fa7d8bb592f6aa6378fb456757354d1441
```


# Overview

Check for segment errors in the [PHOIBLE dev data](https://github.com/phoible/dev).

First, we look at various suggested updates from Cormac Anderson.

He notes:

* Sesotho not in New Guinea ;). PH1017 should be iso: sot, glotto: sout2807 not iso: sso, glotto: siss1243

This is an old issue that's already been resolved [here](https://github.com/phoible/dev/issues/221). 

* In GM 1378 (Mawa) the inventory given lists a number of phonemes, i.e. /ʔ/, /ʔw/, /h/, /f/, /z/ that the source (Roberts 2009) explicitly says are not phonemic.

This issue is raised here:

* https://github.com/phoible/dev/issues/318

Next:

* Would you be happy to represent the Arandic stop-nasal segments in PH 1163, PH 1235, UZ 2158 as pre-stopped nasals, e.g. /ᵗn/, rather than stop-nasal clusters, e.g. /tn/? cf. the prestopped lateral in Kiowa UPSID 409

This will be discussed over email.

Next:

* Just another brief follow up on the Arandic ones. I see now that UZ 2158 uses voiced stop plus nasal, e.g. /bm/, /dn/, whereas PH 1163 and PH 1235 used voiceless stop plus nasal, e.g. /pm/, /tn/. I haven't looked at these languages for a while, so don't know if any phonetic differences are attested in the original material, but I would expect that it is just transcriptional and that there would be a case for using the same convention.

So let's check where we have incompatible transcription practices.

Which rows have voiced/voiceless stop + nasal segments?

```{r}
phoible %>% filter(grepl("bm|pm|dn|tn", Phoneme)) %>% select(InventoryID, Glottocode, LanguageName, SpecificDialect, Phoneme, Allophones, Source) %>% kable()
```

This needs investigation.

Now display all the issues and suggestions that CA made. Some of these also need investigation.

```{r}
cormac <- read_tsv('PHOIBLE data corrections.tsv')
cormac %>% kable()
```



# Older issues

Get a list of unique segments and which sources they appear in.

```{r}
phoneme.source <- phoible %>% select(Phoneme, Source) %>% group_by(Phoneme, Source) %>% distinct()
distinct.segments <- phoneme.source %>% group_by(Phoneme) %>% summarize(Sources=tolower(paste(Source, collapse=";"))) %>% arrange(Phoneme)
expect_equal(nrow(distinct.segments), 3164)
head(distinct.segments)
```

```{r}
# write.csv(distinct.segments, file="distinct-segments.csv", row.names = F)
```

Check for ambigously placed segments, e.g. ring above and below (awful hack here because copy and pasting diacritics in R studio is problematic).

```{r}
rings.below <- distinct.segments %>% filter(grepl("̥", Phoneme))
rings.above <- distinct.segments %>% filter(grepl("̊", Phoneme))
rings <- rbind(rings.above, rings.below)
rings <- rings %>% arrange(Phoneme)
write.csv(rings, file="rings.csv", row.names = F)

kable(rings)
```

Check for ambigously placed segments, e.g. ring above and below (awful hack here because copy and pasting diacritics in R studio is problematic).

```{r}
syllabic.below <- distinct.segments %>% filter(grepl("̩", Phoneme))
syllabic.above <- distinct.segments %>% filter(grepl("̍", Phoneme))
syllabic <- rbind(syllabic.above, syllabic.below)
syllabic <- syllabic %>% arrange(Phoneme)
write.csv(syllabic, file="syllabic.csv", row.names = F)

kable(syllabic)
```

